---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The `anansi` package computes and compares the association between the features of two 'omics datasets that are known to interact based on a database such as KEGG. 

If you use this software, please cite our work.

```{r}
citation("anansi")
```

OK, now let's get started.

```{r, eval = T}
#install and load anansi
#devtools::install_github("thomazbastiaanssen/anansi")
library(anansi)

#load tidyverse, highly recommended for data wrangling 
suppressPackageStartupMessages(library(tidyverse))

#load ggforce for plotting results
library(ggforce)

#load anansi dictionary and human-readable names for KEGG compounds and orthologues
data(dictionary)
data(cpd_translation)
data(KO_translation)

#load example data from FMT Aging study
data(FMT_metadata)
data(FMT_metabs)
data(FMT_KOs)

```

## Data preparation

The main anansi function expects data in the 'anansiWeb' format; Basically a list with exactly three tables: 
The first table should be a count table of metabolites.
The second table should be a count table of functions. 
Both tables should have rows as features and columns as samples.

the third table should be a binary adjacency matrix with the column names of table 1 as rows and the colnames of table 2 as columns. Such a table is provided in the anansi library and is referred to as a dictionary (because you use it to look up which metabolites interact with which functions).  

the weaveWebFromTables() function can be used to parse these tables into 


```{r, eval = T}
#First rid of the "cpd:" part of the names of compounds to match the metabolite table
names(anansi_dic) = gsub(pattern = "cpd:", replacement = "", x = names(anansi_dic))

#Then prune the dictionary list to reduce memory usage
anansi_dic = anansi_dic[names(anansi_dic) %in% rownames(metab)]

#then wrangle the adjacency list to a binary matrix
anansi_dic <- anansi_dic %>% 
  enframe() %>% 
  unnest(cols = c(value)) %>% add_column(present = T) %>%
  pivot_wider(values_from = present, values_fill = F) %>%
  filter(!is.na(value)) %>% 
  column_to_rownames("value") %>% 
  t()

#Clean and CLR-transform the KEGG orthologue table
KOs   <- floor(KOs)
KOs   <- apply(KOs,c(1,2),function(x) as.numeric(as.character(x)))
KOs   <- KOs[apply(KOs == 0, 1, sum) <= (ncol(KOs) * 0.90), ] 

#only keep functions that are represented in the dictionary
KOs   <- KOs[row.names(KOs) %in% colnames(anansi_dic),]

KOs.exp = clr_lite(KOs)

#anansi expects samples to be rows and features to be columns. 
t1 = t(metab)
t2 = t(KOs.exp)
```

## weave a web

```{r, eval = T}
web = weaveWebFromTables(tableY = t1, tableX = t2, dictionary = anansi_dic)
```

## Run anansi

```{r, eval = T}
anansi_out = anansi(web    = web, #generated above
                    method = "pearson", #define the type of correlation used
                    groups = c("A", "B"), #optional, to compare associations between groups
                    adjust.method = "BH", #apply the Benjamini-Hochberg procedure for FDR
                    verbose = T #To let you know what's happening
                    )

```

Anansi gives a nested list of lists as an output. I highly recommend using purrr from the tidyverse to parse it

```{r, eval = T, fig.width = 10, fig.height = 10}
anansi_out %>%
  
  #Induvidually wrangle all results to long format
  map_depth(3, ~data.frame(.x)) %>%
  map_depth(3, ~rownames_to_column(.x, "Compound")) %>%
  map_depth(3, ~pivot_longer(.x, !Compound, names_to = "Function")) %>%
  flatten()  %>%
  
  #Merge estimates, p and q values per model
  map(~reduce(.x, full_join, by = c("Compound", "Function"))) %>%
  map(~rename(.x, c("Estimate" = "value.x", "p.value" = "value.y", "q.value" = "value"))) %>%
  
  #Merge all data to one data.frame
  bind_rows(.id = "Type") %>%
  
  #Organise data by model and parameter type
  pivot_wider(names_from = Type, values_from = c(Estimate, p.value, q.value)) %>%
  
  #Handle correlations separately from total model parameters
  pivot_longer(cols = -c(Compound, Function, 
                         Estimate_modelfit,   p.value_modelfit,   q.value_modelfit, 
                         Estimate_emergent,   p.value_emergent,   q.value_emergent, 
                         Estimate_disjointed, p.value_disjointed, q.value_disjointed)) %>% 
  separate(col = name, into = c("Parameter", "Group"), sep = "_") %>% 
  pivot_wider(names_from = Parameter) %>% 
  #Now it's ready to be piped into ggplot2, though let's clean up a bit more. 
  
  #Only consider canonical interactions
  filter(p.value != 1) %>%
  
  #Only consider interactions where the entire model fits well enough
  filter(q.value_modelfit < 0.1) %>% 
  
  group_by(Compound, Function) %>% 
  filter(min(p.value) < 0.05) %>%
  ungroup() %>%
  
  #Make names of KEGG compounds and functions human-readable
  inner_join(KO_translation, by = "Function") %>%
  inner_join(cpd_translation, by = "Compound") %>%
  
  #Make names pretty
  mutate(Function = paste(Function, KO_name, sep = ": "), 
         Compound = paste(Compound, Compound_name, sep = ": "), 
         Compound = str_remove(Compound, pattern = ";.*")) %>% 
  
  
  #Pipe into ggplot2
  ggplot(aes(x      = Estimate, 
             y      = Function, 
             fill   = Group, 
             alpha  = p.value_disjointed < 0.05)) + 
  
  #Make a vertical dashed red line at x = 0
  geom_vline(xintercept = 0, linetype = "dashed", colour = "red")+
  
  #Points show  raw correlation coefficients
  geom_point(shape = 21, size = 3) + 
  
  #facet per compound
  ggforce::facet_col(~Compound, space = "free", scales = "free_y") + 
  
  #fix the scales, labels, theme and other layout
  scale_y_discrete(limits = rev, position = "right") +
  scale_alpha_manual(values = c("TRUE" = 1, "FALSE" = 1/3)) +
  scale_fill_manual(values = c("Young yFMT" = "#2166ac", 
                               "Aged oFMT"  = "#b2182b", 
                               "Aged yFMT"  = "#ef8a62", 
                               "All"        = "gray"))+
  theme_bw() + 
  ylab("") + 
  xlab("Pearson's rho")
```
