---
title: "Adjacency matrices"
output:
    BiocStyle::html_document:
        toc: true
        toc_float: true
        toc_depth: 2
        number_sections: true
vignette: >
  %\VignetteIndexEntry{2. Working with (bi)adjacency matrices}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
body {
text-align: justify}
</style>

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(anansi)
library(ggplot2)
```

# Overview {-}
A common challenge in integrative analysis of high-dimensional data sets is 
subsequent biological interpretation. Anansi addresses this challenge by only 
considering pairwise associations that are known to occur *a priori*. In order
to achieve this, we need to provide this relational information to anansi in the
form of a (bi)adjacency matrix. 

This vignette is divided by two sections:

* [Section 1](#sec-1) introduces (bi)adjacency matrices in the context of the 
anansi framework.  
* [Section 2](#sec-2) introduces the `anansiWeb` object and demonstrates how to 
work with it

# Understanding adjacency matrices {#sec-1}

## Example from Biology: The Krebs cycle
Recall that the [citric acid cycle, or Krebs cycle](https://en.wikipedia.org/wiki/Citric_acid_cycle),
is a central piece of metabolic machinery. In short, it enables oxidative 
phosphorylation. See figure \@ref(fig:plot-krebs) below for a simplified 
representation. In the figure, there are 8 unique enzymatic reactions and 9 
unique metabolites. 

```{r}
#| plot-krebs, echo=FALSE, fig.small=TRUE,
#| fig.cap="Two simplified representations of the Krebs cycle. Orange marks 
#| metabolites and blue marks enzymes."
library(patchwork)
kd <- data.frame(
    enzyme = c("isocitrate\ndehydrogenase", "aconitase", "aconitase",  
               "citrate\ndehydrogenase", "malate\ndehydrogenase", "fumarase",  
               "succinate\ndehydrogenase","succinyl-CoA\nsynthetase",
               "ketoglutarate\ndehydrogenase"), 
    metab = c("isocitrate","cis-aconitate","citrate","oxaloacetate","L-malate",
              "fumarate","succinate","succinyl-CoA", "ketoglutarate"),
    rad = seq(0, 2*pi * 8/9, length.out = 9)) 
  
ggplot(kd) + 
    annotate("path",
             x=0.75*cos(seq(0,2*pi,length.out=100)),
             y=0.75*sin(seq(0,2*pi,length.out=100))) +
    
    geom_tile(aes(x = 0.8*cos(rad), y =0.8* sin(rad)), width = 0.45, 
              height = 0.15, fill = "orange", color = "black", linewidth = 0.6, 
              show.legend = FALSE) + 
    geom_text(aes(x = 0.8*cos(rad), y =0.8* sin(rad), label = metab), 
              size = 3, show.legend = FALSE) + 
    
    geom_segment(aes(x = 0.75*cos(rad - pi/9), y =0.75* sin(rad - pi/9), 
                     xend = cos(rad - pi/9), yend = sin(rad - pi/9))) +
    
    geom_point(aes(x = 0.75*cos(rad-pi/9), y =0.75* sin(rad-pi/9)), size = 3, 
               fill = "dodgerblue", color = "black", show.legend = FALSE, 
               shape = 21) + 
    geom_text(aes(x = 1.15*cos(rad-pi/9), y =1.15* sin(rad-pi/9),label=enzyme), 
              lineheight = 7/10, size = 3, show.legend = FALSE) + 
    
    scale_y_continuous(limits = c(-1.4, 1.2), expand = c(0, 0)) +
    scale_x_continuous(limits = c(-1.3, 1.3), expand = c(0, 0)) +
    coord_equal(clip = "off") +
    theme_void() 

```

## Adjacency matrices allow us to be specific in our questions

Let's imagine we had (high-dimensional proxy) measurements 
available for both types. For instance, we could have transcription data for the 
enzymes and metabolomics data for the metabolites. We could represent our two 
data sets like two sets of features.

### All vs-all testing 
If we were to perform an all-vs-all integrative analysis between our two data 
sets, comprehensively testing every metabolite-enzyme pair, we'd test 
$8\times9=72$ hypotheses. Figure \@ref(fig:all-v-all) A. 

```{r}
#| all-v-all, echo=FALSE, fig.wide=TRUE, out.width='100%', out.height='60%',
#| fig.cap="Graph representation of possible association sctructures in the 
#| Krebs cycle. **A)** All-vs-all, with a full graph. **B)** Constrained, only 
#| linking metabolites an enzymes that are known to directly interact."


colmat <- data.frame(name = c(
  "aconitase", "citrate\nsynthase", "fumarase", "succinate\ndehydrogenase",
  "isocitrate\ndehydrogenase", "malate\ndehydrogenase", 
  "succinyl-CoA\nsynthetase", "ketoglutarate\ndehydrogenase", "cis-aconitate",
  "citrate", "fumarate", "isocitrate", "L-malate", "oxaloacetate", "succinate", 
  "succinyl-CoA", "ketoglutarate"), 
  y = c(seq(9,1, length.out = 8), 9:1), 
  x = rep(c(0.3, 0.7), times = c(8, 9) ), 
  fill = rep(c("Enzyme", "Metabolite"), times = c(8, 9) ))

p_0 <- ggplot() +
  annotate("segment", x = c(0.28, 0.72), y = 0.875, 
           xend = c(0.28, 0.72), yend = 9.125, color = "gray") + 
  geom_text(data = colmat, aes(y = y, x = x, label = name), colour = "black", 
            size = 3.25, hjust = c(rep(1, 8), rep(0, 9)), lineheight = 7/10, 
            nudge_x = c(rep(-0.03, 8), rep(0.03, 9))) + 
  
  scale_fill_manual(values = c(Metabolite = "orange", Enzyme = "dodgerblue")) +
  coord_cartesian(xlim = c(1/20, 19/20), clip = "off") +
  annotate("text", label = c("Enzymes","Metabolites"), 
           x = c(2/8, 6/8), y = 9.6, size = 4) +  
  ylab(NULL) + xlab(NULL) +
  theme_void()  

p_a <- p_0 + 
  geom_point(data = colmat, aes(y = y, x = x, fill = fill), 
             color = "black", shape = 21, size = 4, show.legend = FALSE)



p_b <- p_0 + geom_segment(data = expand.grid(seq(9,1, length.out = 8), 9:1), 
               aes(y = Var1, yend = Var2), x = 3/10, xend = 7/10, 
               color = "#393D47", inherit.aes = FALSE) + 
  geom_point(data = colmat, aes(y = y, x = x, fill = fill), 
             color = "black", 
             shape = 21, size = 4,  show.legend = FALSE) 
p_c <- p_0 + geom_segment(data = data.frame(
  Var1 = seq(1, 9, length.out = 8)[
    c(8, 8, 7, 6, 5, 8, 4, 6, 3, 7, 3, 5, 2, 2, 1, 4, 1)], 
  Var2 = 
    c(9, 8, 8, 7, 7, 6, 6, 5, 5, 4, 4, 3, 3, 2, 2, 1, 1)), 
  aes(y = Var1, yend = Var2), x = 3/10, xend = 7/10, 
  color = "#393D47", inherit.aes = FALSE) +
  geom_point(data = colmat, aes(y = y, x = x, fill = fill), 
             color = "black", shape = 21, size = 4, show.legend = FALSE) 


(p_b | p_c) + plot_annotation(tag_levels = "A")

```
### Knowledge-informed analyis of specific interactions
Often however, only a subset of these comparisons is scientifically relevant to
investigate. In this case for instance, only 17 associations between 
enzyme-metabolite pairs are likely to make sense. Figure \@ref(fig:all-v-all)
A. Non-selectively testing all 72 associations actively harms statistical power, 
as 55 of these tests likely cannot be interpreted, but their p-values will still
be tallied for the purpose of FDR and FWER. Moreover, they tend to obscure any 
biologically interpretable findings. 

In order to 'know' which features-pairs in your data set should be considered, 
anansi requires a *biadjacency matrix*. The biadjacency matrix corresponding to
to the sparse graph in figure \@ref(fig:all-v-all) B can be seen in figure 
\@ref(fig:krebs-biadj). 

```{r}
#| krebs-biadj, echo=FALSE, fig.small=TRUE,
#| fig.cap="Matrix representation of the constrained analysis of the Krebs cycle, 
#| with cells marked with 'X' signifying tested feature-pairs."

kd$lab <- "X"
ggplot(kd) +
  geom_text(aes(label = lab), x = I(0.5), y = I(0.5)) +
  facet_grid(enzyme ~ metab , switch = "y", space = "free_x") + 
  theme_bw() +
  theme(strip.text.y.left = element_text(
    angle = 0, size = 10, lineheight = 7/10, hjust = 1),
        strip.text.x.top = element_text(angle = 90, size = 10, hjust = 0),
        strip.background.x = element_rect(fill = "orange", color = "black"), 
        strip.background.y = element_rect(fill = "dodgerblue", color = "black"), 
        panel.spacing = unit(0.1, "lines")) +
  labs(x = NULL, y = NULL)

```

# General use in anansi {#sec-2}
In anansi, we work with biadjacency matrices in the `anansiWeb` S4 class, which 
consists of a biadjacency matrix, in this case we call it a `dictionary`, as 
well as two tables of observations that should be analysed jointly. 

The anansi package offers some functions that should be sufficient for basic 
analysis. To enable more advanced users to pursue non-standard applications, our
methodology is compatible with several popular interfaces. 

## Make a biadjacency matrix with `weaveWeb()`
The recommended default interface to generate an `anansiWeb` object is 
`weaveWeb()`. Besides a traditional interface, it also accepts R `formula` 
syntax. Once we have an `anansiWeb` object, we can use the $ operator to access 
the biadjacency matrix in the 'dictionary' slot.

```{r weaveWeb-intro}
# Two interfaces
form.web <- weaveWeb(ec ~ ko)

trad.web <- weaveWeb(y = "ec", x = "ko")

# Same output
identical(form.web, trad.web)

# Get the biadjacency matrix using the $ operator
head( form.web$dictionary,  c(10, 10) )
```
Note that the majority of this matrix consists of dots, it's a sparse matrix, 
which is a format that can considerably speed up calculations involving huge, 
mostly empty matrices. For further reading, see the 
[Matrix package website](https://matrix.r-forge.r-project.org/). 

### `weaveWeb()` input: `link`
`weaveWeb()` takes additional arguments, let's focus on `link`, which will take 
the information necessary to link features between input data sets. The anansi 
package includes such data for the [KEGG database](https://www.genome.jp/kegg/), 
let's take a look at some of it:

```{r peakec2ko}
head(ec2ko)
```

### Structure of `link` input
The format of the `data.frame` is important: It should have two named columns. 
Each column contains feature IDs of one type (in this case, `ec` and `ko`, which
is how the KEGG database refers to 
[Enzyme Commission numbers](https://enzyme.expasy.org/) and [kegg orthologes](https://www.genome.jp/kegg/ko.html), respectively. If two feature 
IDs are on the same row of the `data.frame`, it means those features are paired
(hence adjacency). 

### linking between terms
The column names correspond to the types of feature IDs that should be linked. 
These names are used throughout the `anansiWeb` workflow. They can be directly 
called from an `anansiWeb` object:

```{r terms}
# Note that the formula controls the order 
names(dimnames(
  weaveWeb(ko~ec)$dictionary
  ))

# terms() is short for the above
terms(weaveWeb(ec~ko))
```
## The pre-packaged kegg linking map
By default, `weaveWeb()` will assume to use the pre-packaged kegg link map, 
which consists of two similarly structured `data.frame`s in a list, one of which
we just inspected. We can call it directly using `kegg_link()`: 
```{r kegg_web}
# Only print the first few entries
lapply( kegg_link(), head)

```

### linking across two `data.frame`
Note the column names in the two `data.frames`: `ec`, `cpd` and `ko`. Two of 
these correspond to the formula we used just used: `weaveWeb(ec ~ ko)`. We 
didn't use the term `cpd`. Further, `ec` is present in both `data.frame`s. 

We can use `weaveWeb()` to make a biadjacency matrix between any combination of 
one or two similarly structured `data.frame`s, presented as a list. For the pre-
packaged data set, this means we can link between any two of `ec`, `cpd` and 
`ko`, in either order. 

```{r web_formula}
# Formula notation
head( weaveWeb(cpd ~ ko)$dictionary, c(10, 10) )

# Character notation
head( weaveWeb(y = "ko", x = "cpd")$dictionary, c(10, 10) )
```

## Use custom biadjacency matrices with as_web()
The most flexible way to make an `anansiWeb` object is through `as_web()`: 
```{r as_web}
# Prep some dummy input tables
  tX <- matrix(rnorm(30), nrow = 6)
  tY <- matrix(rnorm(42), nrow = 6)

# Prep biadjacency matrix    
# base::matrix is fine too, but will get coerced to Matrix::Matrix.
  d <- matrix(
    data = sample(x = c(TRUE, FALSE), size = 35, replace = TRUE), 
    nrow = NCOL(tY), 
    ncol = NCOL(tX)
    )
  
# make the anansiWeb
 w <- as_web(
    tableY = tY, 
    tableX = tX,
    dictionary = d
  )

# Confirm the warning:  
  terms(w)
```
## Additional approaches
There are numerous ways in which we can define an adjacency matrix. Here, we 
demonstrate a graph-based and matrix based approach.

```{r init-list, echo=FALSE}
krebs_edge_df <- data.frame(
 Enzyme = c("aconitase", "aconitase", "citrate synthase", 
"fumarase", "succinate dehydrogenase", "aconitase", "isocitrate dehydrogenase", 
"fumarase", "malate dehydrogenase", "citrate synthase", "malate dehydrogenase", 
"succinate dehydrogenase", "succinyl-CoA synthetase", "succinyl-CoA synthetase", 
"ketoglutarate dehydrogenase", "isocitrate dehydrogenase", 
"ketoglutarate dehydrogenase"), 
 Metabolite = c("cis-aconitate", 
"citrate", "citrate", "fumarate", "fumarate", "isocitrate", "isocitrate", 
"L-malate", "L-malate", "oxaloacetate", "oxaloacetate", "succinate", 
"succinate", "succinyl-CoA", "succinyl-CoA", 
"ketoglutarate", "ketoglutarate"))
```

### adjacency matrices with `igraph`
Importantly, (bi)adjacency matrices can be understood as graphs. Two common 
packages that deal with graphs are [`igraph`](https://r.igraph.org/) and [`graph`](https://bioconductor.org/packages/graph/). 
```{r igraph-1, message=FALSE}
library(igraph)

# Convert data.frame to graph
g <- graph_from_data_frame(krebs_edge_df, directed = FALSE)

```

Now that we have constructed a graph, we still need to identify which features, 
vertices, belong to which data modality, in this case either enzymes and 
metabolites. `igraph` does this using a boolean vector, where `TRUE` values 
become columns, which corresponds to `tableX`, whereas `FALSE` values become 
rows, corresponding to `tableY`.

```{r igraph-2, message=FALSE}
V(g)$type <- V(g)$name %in% krebs_edge_df$Enzyme

# Now that we've defined type, we can convert to biadjacency matrix: 
bi_mat1 <- as_biadjacency_matrix(g, sparse = TRUE)

head(bi_mat1, n = c(4, 5))
```
Though biadjacency support in `graph` is currently limited, We note that igraph 
and graph objects can be converted using the `graph_from_graphnel()` and 
`as_graphnel()` functions in `igraph`.

### adjacency matrices with `Matrix`
We can also define a matrix directly. Conveniently, sparse matrices can be 
defined easily from our starting data. The `Matrix` library provides fantastic 
support for this.
```{r Matrix, message=FALSE}
library(Matrix)

# For this approach, it's useful to prepare the input as two factors:
enzymes     <- factor(krebs_edge_df$Enzyme)
metabolites <- factor(krebs_edge_df$Metabolite)

# We can get integers out of factors, corresponding to their level indices 
bi_mat2 <- sparseMatrix(i = as.integer(enzymes), 
             j = as.integer(metabolites), 
             dimnames = list(
               levels(enzymes),
               levels(metabolites)
               )
             )

head(bi_mat2, n = c(4, 5))

```

# Session info

```{r}
sessionInfo()
```
