---
title: "Getting started with anansi"
output:
    BiocStyle::html_document:
        toc: true
        toc_float: true
        toc_depth: 2
        number_sections: true
vignette: >
  %\VignetteIndexEntry{1. Getting started with anansi}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
body {
text-align: justify}
</style>

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      tidy.opts = list(width.cutoff = 80), 
                      tidy = TRUE)
# long links
adj_link <- paste0("https://thomazbastiaanssen.github.io/anansi/", 
               "articles/adjacency_matrices.html")
dif_link <- paste0("https://thomazbastiaanssen.github.io/anansi/", 
               "articles/differential_associations.html")

library(anansi)
```

# Introduction
The `anansi` package computes and compares the association between the features 
of two 'omics datasets that are known to interact based on a database such as 
KEGG. Studies including both functional microbiome and metabolomics data are
becoming more common. Often, it would be helpful to integrate both datasets in
order to see if they corroborate each others patterns. However, all-vs-all
association analyses are imprecise and likely to yield spurious associations.
This package takes a knowledge-based approach to constrain association search
space, only considering  metabolite-function interactions that have been
recorded in a pathway database. In addition, it provides a framework to assess
differential associations. 

While `anansi` is geared towards metabolite-function interactions in the context 
of host-microbe interactions, it is perfectly capable of handling any other pair 
of data sets where some features interact canonically. 

## Installation
Install the development version from
[GitHub](https://github.com/thomazbastiaanssen/anansi) with `remotes`: 
```{r, 'remotes-install', eval = FALSE}
install.packages("remotes")
remotes::install_github("thomazbastiaanssen/anansi")
```

## Setup
```{r, eval=TRUE}
# load anansi
library(anansi)

# load ggplot2 and ggforce to plot results
library(ggplot2)
library(ggforce)

# load example data + metadata from FMT Aging study
data(FMT_data)
```

# Data preparation

The main `anansi` function expects data in the `AnansiWeb` class format, which 
contains three tables: 
The first two tables, `tableY` and `tableX` are the feature tables of your two
data modalities (e.g., metabolites, genes, microbes, ...). 
Both tables should have columns as features and rows as samples.

The third table should be a binary adjacency matrix with the column names of 
`tableY` as rows and the column names of `tableX` as columns. An example of such
an adjacency matrix comes with the `anansi` package and is referred to as a
dictionary (because it is used to look up which metabolites interact with which
functions).  

Though most of the examples use metabolites and functions, `anansi` is able to 
handle any type of big/'omics data, as long as there is a dictionary available. 
Because of this, anansi uses the type-naive nomenclature `tableY` and `tableX`. 
The Y and X refer to the position these measurements will have in the linear 
modeling framework: 

$$
Y \sim X \times{covariates}
$$

### A note on functional microbiome data

Two common questions in the host-microbiome field are "Who's there?" and "What 
are they doing?". Techniques like 16S rRNA sequencing and shotgun metagenomics 
sequencing are most commonly used to answer the first question. The second
question can be a bit more tricky, as it often requires functional inference 
software to address them. For 16S sequencing, algorithms like PICRUSt2 and
Piphillin can be used to infer function. For shotgun metagenomics, HUMANn3 in
the bioBakery suite can be used. All of these algorithms can produce functional
count data in terms of KEGG Orthologues (KOs) organised as tables, which can be
directly passed to `anansi` as follows: 

```{r, eval=TRUE}
# anansi expects samples to be rows and features to be columns.
t1 <- t(FMT_metab)
t2 <- t(FMT_KOs)
```

# Weave a web🕸️

The `weaveWeb()` function can be used to parse the tables `t1` and `t2` that 
we prepared above into an `AnansiWeb` object. The `AnansiWeb` format is a 
necessary input file for the main `anansi` workflow. It allows `anansi` to keep 
track of which features from the two input data sets should be considered
as pairs. 

`anansi` provides a pre-built map between ko, cpd and ec feature annotations
from the KEGG database, but users can also provide their own custom maps.
[See vignette on adjacency matrices.]`r sprintf("(%s)", adj_link)`

```{r, eval=TRUE}
# Create AnansiWeb object from tables
web <- weaveWeb(
    cpd ~ ko,
    tableY = t1,
    tableX = t2,
    link = kegg_link()
)
```

# Run anansi🕷️

The main workspider is called `anansi` just like the package. Generally, three
main arguments should be specified:

1. `web`: an `AnansiWeb` object, such as the one we generated in the above step.
2. `formula`: a linking formula. For instance, to assess differential
   associations between treatments, we use the formula `~ Legend`.
3. `metadata`: a `data.frame` with additional information about the data set.
   Its rows correspond to samples and its columns to variables that can be used
   in `formula` as c.

```{r, eval=TRUE}
# Run anansi on AnansiWeb object
anansi_out <- anansi(
    web = web,
    formula = ~ Legend,
    metadata = FMT_metadata,
    adjust.method = "BH",
    verbose = TRUE
)
```
# Reporting output📝

By default, `anansi` outputs a table in the `wide` format. For general 
reporting, we  recommend sticking to the more legible `table` format by 
specifying the `return.format` argument. Let's take a look at the structure of 
the output table:

```{r, eval=TRUE}
# General structure of anansi() output
head(anansi_out, c(5, 5))

# Let's take a look at the column names
colnames(anansi_out)
```

## Output statistics
We can see that the output of `anansi()` is structured so that there is one
feature pair per row, marked by the first two columns. For each of those feature
pairs, the table contains outcome parameters from several models. In order, we
have pooled correlations, group-wise correlations, full model, disjointed and
emergent associations. For each of these models we report four parameters. 
For correlations we provide the Pearson's coefficient ($\rho$) and the
$t$-statistic, whereas we provide their squared counterparts for the linear
regression models: $R^2$ and the $F$ statistic. For all models, we additionally
provide p-values and adjusted p-values. For more on this, 
[see the vignette on differential associations]`r sprintf("(%s)", dif_link)`.

| Association | Method            | Statistics  | Significance  |
|-------------|-------------------|-------------|---------------|
| Pooled      | Correlation       | $\rho$, $t$ | p, adjusted p |
| Group-wise  | Correlation       | $\rho$, $t$ | p, adjusted p |
| Full        | Linear regression | $R^2$, $F$  | p, adjusted p |
| Disjointed  | Linear regression | $R^2$, $F$  | p, adjusted p |
| Emergent    | Linear regression | $R^2$, $F$  | p, adjusted p |

# Plot the results

The long format can be helpful to plug the data into `ggplot2`. Here, we 
recreate the style of the figure from the FMT Aging study.

```{r}
#| plot_FMT, eval=TRUE, fig.width=10, fig.height=8, tidy=FALSE,
#| fig.alt="Differential association plot, with facets separating metabolites
#| and metabolites on the y-axis. x-axis depicts pearson's correlation
#| coefficient."
# Use tidyr to wrangle the correlation r-values to a single column
library(tidyr)

anansiLong <- anansi_out |>
    pivot_longer(starts_with("All") | contains("FMT")) |>
    separate_wider_delim(
        name,
        delim = "_", names = c("cor_group", "param")
    ) |>
    pivot_wider(names_from = param, values_from = value)

# Now it's ready to be plugged into ggplot2, though let's clean up a bit more.
# Only consider interactions where the entire model fits well enough.
anansiLong <- anansiLong[anansiLong$full_q.values < 0.2, ]

ggplot(anansiLong) +

    # Define aesthetics
    aes(
        x = r.values, 
        y = feature_X,
        fill = cor_group, 
        alpha = disjointed_Legend_p.values < 0.05
        ) +

    # Make a vertical dashed red line at x = 0
    geom_vline(xintercept = 0, linetype = "dashed", colour = "red") +

    # Points show  raw correlation coefficients
    geom_point(shape = 21, size = 3) +

    # facet per compound
    ggforce::facet_col(~feature_Y, space = "free", scales = "free_y") +

    # fix the scales, labels, theme and other layout
    scale_y_discrete(limits = rev, position = "right") +
    scale_alpha_manual(
        values = c("TRUE" = 1, "FALSE" = 1 / 3),
        "Disjointed association\np < 0.05"
    ) +
    scale_fill_manual(
        values = c(
            "Young yFMT" = "#2166ac",
            "Aged oFMT" = "#b2182b",
            "Aged yFMT" = "#ef8a62",
            "All" = "gray"
        ),
        breaks = c("Young yFMT", "Aged oFMT", "Aged yFMT", "All"),
        name = "Treatment"
    ) +
    theme_bw() +
    ylab("") +
    xlab("Pearson's \u03c1")
```

## Plotting multiple feature pairs at once with patchwork
We can also also use the `AnansiWeb` object to pull up specific pairs of
features. The [`patchwork`](https://patchwork.data-imaginist.com/) provides a
wonderful interface to combine plots. We can use `getFeaturePairs()` to
extract the relevant data from our `AnansiWeb` object in a convenient list
format:   
```{r}
#| multiplot, eval=TRUE, message=FALSE, tidy = FALSE
library(patchwork)

# Get a list containing data.frames for each feature pair.
feature_pairs <- getFeaturePairs(web)

# For each, prepare a ggplot object with those features on x & y axes.
plot_list <- lapply(
    feature_pairs,
    FUN = function(pair_i) {
        ggplot(pair_i) +
            aes(
                y = .data[[colnames(pair_i)[1L]]],
                x = .data[[colnames(pair_i)[2L]]]
            ) +
            # We'll use facet_grid for labels instead of labs().
            facet_grid(colnames(pair_i)[1L] ~ colnames(pair_i)[2L])
    }
)

# Feed the first six of our ggplots to patchwork, using wrap_plots()
wrap_plots(plot_list[seq_len(6L)], guides = "collect") &
    # Continue as with an ordinary ggplot, but use & instead of +
    geom_point(
        shape = 21, show.legend = FALSE,
        aes(fill = FMT_metadata$Legend)
    ) &
    geom_smooth(
        method = "lm", se = FALSE,
        aes(colour = FMT_metadata$Legend),
        show.legend = FALSE
    ) &

    # Appearance
    scale_fill_manual(
        aesthetics = c("colour", "fill"),
        values = c(
            "Young yFMT" = "#2166ac",
            "Aged oFMT" = "#b2182b",
            "Aged yFMT" = "#ef8a62",
            "All" = "gray"
        )
    ) &
    theme_bw() &
    labs(x = NULL, y = NULL)
```
These types of visualizations of feature pairs work best when the amount of 
feature pairs to be plotted is modest.

### Note on using `patchwork` with `anansi`: 
Especially for large numbers of feature pairs, it's much more efficient to add 
shared elements to the patchwork using the `&` operator, rather than adding them
to the initial `ggplot()` calls in the `lapply()` loop. 


# Bioconductor pipeline

`anansi` also supports the Bioconductor class `MultiAssayExperiment` (MAE) as an
alternative to `AnansiWeb`. The latter can be converted to a MAE with the
function `asMAE`.

```{r}
#| bioc1, eval=TRUE, message=FALSE
library(MultiAssayExperiment)

# Store sample metadata into metadata slot of web object
web@metadata <- `rownames<-`(FMT_metadata, FMT_metadata$Sample_ID)

# Convert web to mae
mae <- asMAE(web)
```

The MAE object can be passed to the `weaveWeb` function, which returns the same
output, ready for the `anansi` function.

```{r}
#| bioc2, eval=TRUE

web2 <- weaveWeb(
    mae,
    tableY = "cpd", tableX = "ko",
    link = kegg_link()
)

out <- anansi(
    web = web2,
    formula = ~Legend,
    adjust.method = "BH",
    verbose = TRUE, 
)
```

Finally, the output of `anansi` can be visualised with `plotAnansi` by
specifying the type of association to use.

```{r}
#| bioc3, eval=TRUE, fig.width = 10, fig.height = 8

# Filter associations by q value
out <- out[out$full_q.values < 0.2, ]

# Visualize disjointed associations
plotAnansi(out,
    association.type = "disjointed",
    model.var = "Legend",
    signif.threshold = 0.05,
    fill_by = "group"
)
```

# Session info

```{r}
sessionInfo()
```
