---
title: "Getting started with anansi"
output:
    BiocStyle::html_document:
        toc: true
        toc_float: true
        toc_depth: 2
        number_sections: true
vignette: >
  %\VignetteIndexEntry{1. Getting started with anansi}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
body {
text-align: justify}
</style>

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    tidy.opts = list(width.cutoff = 80),
    tidy = TRUE
)
# long links
adj_link <- paste0(
    "https://thomazbastiaanssen.github.io/anansi/",
    "articles/adjacency_matrices.html"
)
dif_link <- paste0(
    "https://thomazbastiaanssen.github.io/anansi/",
    "articles/differential_associations.html"
)

library(anansi)
```
# Overview {-}

This vignette can be divided into four sections:

* [Section 1](#sec-anansi-1) introduces the `anansi` framework.
* [Section 2](#sec-anansi-2) goes through a simple workflow and required input.
* [Section 3](#sec-anansi-3) discusses more advanced applications of anansi. 

# Introduction {#sec-anansi-1}
The `anansi` package computes and compares the association between the features 
of two 'omics datasets that are known to interact based on a database such as 
KEGG. Studies including both functional microbiome and metabolomics data are
becoming more common. Often, it would be helpful to integrate both datasets in
order to see if they corroborate each others patterns. However, all-vs-all
association analyses are imprecise and likely to yield spurious associations.
This package takes a knowledge-based approach to constrain association search
space, only considering  metabolite-function interactions that have been
recorded in a pathway database. In addition, it provides a framework to assess
differential associations. 

While `anansi` is geared towards metabolite-function interactions in the context 
of host-microbe interactions, it is perfectly capable of handling any other pair 
of data sets where some features interact canonically. 

## A note on functional microbiome data

Two common questions in the host-microbiome field are "Who's there?" and "What 
are they doing?". Techniques like 16S rRNA sequencing and shotgun metagenomics 
sequencing are most commonly used to answer the first question. The second
question can be a bit more tricky, as it often requires functional inference 
software to address them. For 16S sequencing, programs like 
[PICRUSt2](https://github.com/picrust/picrust2/) can be used to infer function. 
For shotgun metagenomics, this can be done with programs such as 
[HUMANn3 in the bioBakery suite](https://github.com/picrust/picrust2) and [woltka](https://github.com/qiyunzhu/woltka). All of these algorithms can 
produce functional count data in terms of KEGG Orthologues (KOs) organised as 
tables, which can be directly passed to `anansi`. 

# Getting started with anansi {#sec-anansi-2}

## Installation
Install the development version from
[GitHub](https://github.com/thomazbastiaanssen/anansi) with `remotes`: 
```{r, 'remotes-install', eval = FALSE}
install.packages("remotes")
remotes::install_github("thomazbastiaanssen/anansi")
```

## Setup 
```{r, eval=TRUE}
# load anansi
library(anansi)

# load ggplot2 and ggforce to plot results
library(ggplot2)
library(ggforce)

# load example data + metadata from FMT Aging study
data(FMT_data)
```
    
## Data preparation

The anansi framework requires input data to be prepared in a specific format, 
namely the `AnansiWeb` S7 class. This is to ensure that features are correctly 
linked across input datasets. We will first discuss the required input data and 
then demonstrate the `weaveWeb` method, which is the recommended way to prepare 
input for the anansi framework. 

## Input formatting for anansi

The main `anansi` function expects data in the `AnansiWeb` class format, which 
contains four tables: 

### feature tables: `tableY` and `tableX`. 

The first two tables, `tableY` and `tableX` are the feature tables of your two
data modalities (e.g., metabolites, genes, microbes, ...). Both tables should 
have columns as features and rows as samples. Below is an example of an 
appropriately formatted feature table, we will use it in the sections below. 

```{r ko-table-format, echo = FALSE}
cat("anansi expects samples to be rows and features to be columns:\n")
t(FMT_KOs[seq(4L),seq(10L)])
```

```{r, eval=TRUE}
# Transpose demo tables to input format
tableY <- t(FMT_metab)
tableX <- t(FMT_KOs)
```

### linking information: `link`.

The third table, `link`, provides the information to link the features across 
the first two tables, `tableY` and `tableX`. `anansi` provides a pre-built map 
between ko, cpd and ec feature annotations from the KEGG database, but users can
also provide their custom mapping information. We will not expand on here, but
[see vignette on adjacency matrices.]`r sprintf("(%s)", adj_link)`
Also see `?anansi::MultiFactor()` 

### Additional sample data for analysis: Metadata

The final table is `metadata`: a `data.frame` with additional information about 
the data set. Its rows correspond to samples and its columns to variables that 
can be used as covariates in downstream `anansi()` analysis.

```{r metadata-format, echo = FALSE}
cat("anansi expects samples to be rows and features to be columns:\n")
`rownames<-`(FMT_metadata, NULL)[seq(10L),]
```

## Weave a web🕸️

The `weaveWeb()` function accepts the four input tables discussed above and 
collates them into an `AnansiWeb` object. The `AnansiWeb` format is a necessary 
input file for the main `anansi` workflow. It allows `anansi` to keep track of 
which features from the two input data sets should be considered as pairs. 

### Specify features to link

Additionally, `weaveWeb()` requires the user to specify the names of the feature
types that should be linked. These names need to be present as the columns of 
the `link` table. In this case, we'll link KEGG compound (cpd) to KEGG 
Orthologues (ko). 

```{r, eval=TRUE}
# Create AnansiWeb object from tables
web <- weaveWeb(
    x = "ko", y = "cpd", 
    link = kegg_link(),
    tableY = tableY,
    tableX = tableX,
    metadata = FMT_metadata
)
```

### On model notation and the names `tableY` and `tableX` 

Though most of the examples use metabolites and functions, `anansi` is able to 
handle any type of big/'omics data, as long as there is a dictionary available. 
Because of this, anansi uses the type-naive nomenclature `tableY` and `tableX`. 
The Y and X refer to the position these measurements will have in the linear 
modeling framework: 

$$
Y \sim X \times{covariates}
$$
### `weaveWeb` with Bioconductor pipeline

As an alternative to providing input as raw tables, `weaveWeb()` also accepts 
the Bioconductor classes `MultiAssayExperiment` (MAE) and 
`TreeSummarizedExperiment` (TSE) as input. Conversely, an `AnansiWeb` object can
be converted to a MAE or TSE with the functions `asMAE()` and `asTSE()`.

```{r}
#| bioc1, eval=TRUE, message=FALSE
# Convert web to mae
mae <- asMAE(web)
```

The MAE object can be passed to the `weaveWeb()` function, which returns the same
output, ready for the `anansi` function.

```{r}
#| bioc2, eval=TRUE

web.mae <- weaveWeb(
    mae,
    tableY = "cpd", tableX = "ko",
    link = kegg_link()
)

out.mae <- anansi(
    web = web.mae,
    formula = ~Legend,
    adjust.method = "BH",
    verbose = TRUE
)

identical(web, web.mae)
```


## Run anansi🕷️

The main workspider is called `anansi` just like the package. Generally, two
main arguments should be specified:

1. `web`: an `AnansiWeb` object, such as the one we generated in the above step.
2. `formula`: a formula specifying which associations to test. For instance, to 
assess differential associations between treatments, we use the formula 
`~ Legend` (corresponding to the `Legend` column in the `metadata` table we 
provided to `weaveWeb()` above).

```{r, eval=TRUE}
# Run anansi on AnansiWeb object
anansi_out <- anansi(
    web = web,
    formula = ~Legend,
    adjust.method = "BH",
    verbose = TRUE
)
```
## Reporting output📝

By default, `anansi` outputs a table in the `wide` format. For general 
reporting, we  recommend sticking to the more legible `table` format by 
specifying the `return.format` argument. Let's take a look at the structure of 
the output table:

```{r, eval=TRUE}
# General structure of anansi() output
head(anansi_out, c(5, 5))

# Let's take a look at the column names
colnames(anansi_out)
```

### Output statistics
We can see that the output of `anansi()` is structured so that there is one
feature pair per row, marked by the first two columns. For each of those feature
pairs, the table contains outcome parameters from several models. In order, we
have pooled correlations, group-wise correlations, full model, disjointed and
emergent associations. For each of these models we report four parameters. 
For correlations we provide the Pearson's coefficient ($\rho$) and the
$t$-statistic, whereas we provide their squared counterparts for the linear
regression models: $R^2$ and the $F$ statistic. For all models, we additionally
provide p-values and adjusted p-values. For more on this, 
[see the vignette on differential associations]`r sprintf("(%s)", dif_link)`.

| Association | Method            | Statistics  | Significance  |
|-------------|-------------------|-------------|---------------|
| Pooled      | Correlation       | $\rho$, $t$ | p, adjusted p |
| Group-wise  | Correlation       | $\rho$, $t$ | p, adjusted p |
| Full        | Linear regression | $R^2$, $F$  | p, adjusted p |
| Disjointed  | Linear regression | $R^2$, $F$  | p, adjusted p |
| Emergent    | Linear regression | $R^2$, $F$  | p, adjusted p |

## Plot the results

Finally, the output of `anansi` can be visualized with `plotAnansi` by
specifying the type of association to use.

```{r}
#| bioc3, eval=TRUE, fig.width = 10, fig.height = 8

# Filter associations by q value
anansi_out <- anansi_out[anansi_out$full_q.values < 0.2, ]

# Visualize disjointed associations
plotAnansi(
    anansi_out,
    association.type = "disjointed",
    model.var = "Legend",
    signif.threshold = 0.05,
    fill_by = "group"
)
```

# Advanced applications and customization {#sec-anansi-3}

## Apply user-defined function on each pair of features

A typical `anansi` workflow, as the one above, estimates associations between 
each pair of features. More generally, we can use `pairwiseApply()` on the same 
`AnansiWeb` input object to apply a user-defined function on each pair of 
features. The function passes as the `FUN` argument is applied on each pair of 
features, which are pre-mapped to `y` and `x`:  `function(x, y)`. 
[Also see vignette on adjacency matrices.]`r sprintf("(%s)", adj_link)`

```{r, pairwiseApply1}
# For each feature pair, was the value for x higher than the value for y?
pairwise_gt <- pairwiseApply(
    X = web,
    FUN = function(x, y) x > y, 
    USE.NAMES = TRUE
)

pairwise_gt[1:5, 1:5]
```


```{r, pairwiseApply2}
# Run cor() on each pair of features
pairwise_cor <- pairwiseApply(
    X = web,
    FUN = function(x, y) cor(x, y), 
    USE.NAMES = FALSE
)

pairwise_cor
```

## Plotting examples

Automatic plotting functions such as `plotAnansi()` provide a convenient 
and code-light solution to visualization. However, some may prefer interfacing 
directly with the `ggplot()` code. We provide code to recreate the above figure:

```{r}
#| plot_FMT, eval=TRUE, fig.width=10, fig.height=8, tidy=FALSE,
#| fig.alt="Differential association plot, with facets separating metabolites
#| and metabolites on the y-axis. x-axis depicts pearson's correlation
#| coefficient."
# Use tidyr to wrangle the correlation r-values to a single column
library(tidyr)

anansiLong <- anansi_out |>
    pivot_longer(starts_with("All") | contains("FMT")) |>
    separate_wider_delim(
        name,
        delim = "_", names = c("cor_group", "param")
    ) |>
    pivot_wider(names_from = param, values_from = value)

# Now it's ready to be plugged into ggplot2, though let's clean up a bit more.
# Only consider interactions where the entire model fits well enough.
anansiLong <- anansiLong[anansiLong$full_q.values < 0.2, ]

ggplot(anansiLong) +

    # Define aesthetics
    aes(
        x = r.values,
        y = feature_X,
        fill = cor_group,
        alpha = disjointed_Legend_p.values < 0.05
    ) +

    # Make a vertical dashed red line at x = 0
    geom_vline(xintercept = 0, linetype = "dashed", colour = "red") +

    # Points show  raw correlation coefficients
    geom_point(shape = 21, size = 3) +

    # facet per compound
    ggforce::facet_col(~feature_Y, space = "free", scales = "free_y") +

    # fix the scales, labels, theme and other layout
    scale_y_discrete(limits = rev, position = "right") +
    scale_alpha_manual(
        values = c("TRUE" = 1, "FALSE" = 1 / 3),
        "Disjointed association\np < 0.05"
    ) +
    scale_fill_manual(
        values = c(
            "Young yFMT" = "#2166ac",
            "Aged oFMT" = "#b2182b",
            "Aged yFMT" = "#ef8a62",
            "All" = "gray"
        ),
        breaks = c("Young yFMT", "Aged oFMT", "Aged yFMT", "All"),
        name = "Treatment"
    ) +
    theme_bw() +
    ylab("") +
    xlab("Pearson's \u03c1")
```

## Plotting multiple feature pairs at once with patchwork

We can also also use the `AnansiWeb` object to pull up specific pairs of
features. The [`patchwork`](https://patchwork.data-imaginist.com/) provides a
wonderful interface to combine plots. We can use `getFeaturePairs()` to
extract the relevant data from our `AnansiWeb` object in a convenient list
format:   
```{r}
#| multiplot, eval=TRUE, message=FALSE, tidy = FALSE
library(patchwork)

# Get a list containing data.frames for each feature pair.
feature_pairs <- getFeaturePairs(web)

# For each, prepare a ggplot object with those features on x & y axes.
plot_list <- lapply(
    feature_pairs,
    FUN = function(pair_i) {
        ggplot(pair_i) +
            aes(
                y = .data[[colnames(pair_i)[1L]]],
                x = .data[[colnames(pair_i)[2L]]]
            ) +
            # We'll use facet_grid for labels instead of labs().
            facet_grid(colnames(pair_i)[1L] ~ colnames(pair_i)[2L])
    }
)

# Feed the first six of our ggplots to patchwork, using wrap_plots()
wrap_plots(plot_list[seq_len(6L)], guides = "collect") &
    # Continue as with an ordinary ggplot, but use & instead of +
    geom_point(
        shape = 21, show.legend = FALSE,
        aes(fill = FMT_metadata$Legend)
    ) &
    geom_smooth(
        method = "lm", se = FALSE,
        aes(colour = FMT_metadata$Legend),
        show.legend = FALSE
    ) &

    # Appearance
    scale_fill_manual(
        aesthetics = c("colour", "fill"),
        values = c(
            "Young yFMT" = "#2166ac",
            "Aged oFMT" = "#b2182b",
            "Aged yFMT" = "#ef8a62",
            "All" = "gray"
        )
    ) &
    theme_bw() &
    labs(x = NULL, y = NULL)
```
These types of visualizations of feature pairs work best when the amount of 
feature pairs to be plotted is modest.

### Note on using `patchwork` with `anansi`: 

Especially for large numbers of feature pairs, it's much more efficient to add 
shared elements to the patchwork using the `&` operator, rather than adding them
to the initial `ggplot()` calls in the `lapply()` loop. 


# Session info

```{r}
sessionInfo()
```
