---
title: "Association testing"
output:
    BiocStyle::html_document:
        toc: true
        toc_float: true
        toc_depth: 2
        number_sections: true
vignette: >
  %\VignetteIndexEntry{3. Differential associations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
body {
text-align: justify}
</style>

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(anansi)
```


# Overview {-}
A common challenge in integrative analysis of high-dimensional data sets is 
subsequent biological interpretation. Anansi addresses this challenge by only 
considering pairwise associations that are known to occur *a priori*. In order
to achieve this, we need to provide this relational information to anansi in the
form of a (bi)adjacency matrix. 

This vignette is divided by three sections:

* [Section 1](#sec-da-1) introduces the anansi framework and covers the *full 
model*.  
* [Section 2](#sec-da-2) Covers differential associations
* [Section 3](#sec-da-3) Discusses incorporating repeated measures through 
`Error()` notation. 

# Getting started: The full model {#sec-da-1}
After determining which pairs of features should be investigated, `anansi` fits
several linear models for which it returns summary statistics. These user has 
control over these models through the `formula` argument in `anansi(). `

### Setup 
[See the vignette](https://thomazbastiaanssen.github.io/anansi/articles/adjacency_matrices.html) for more about adjacency matrices and `weaveWeb()`. 
```{r formula-web-full} 
# Setup AnansiWeb with some dummy data
web <- krebsDemoWeb()

```
## Formula syntax
Anansi supports arbitrarily complex linear models as well as longitudinal 
models sing the R formula syntax. The most basic model in anansi would be 
`y ~ x`, where `y` is the feature from `tableY` and `x` is the corresponding 
feature in `tableX`. 
However, these two variables shouldn't be mentioned in the model. Rather, only additional covariates that could interact with x should be provided in the 
`formula` argument. We also see this back in the message: 

```{r formula-anansi-full} 
out <- anansi(web,
    formula = ~ group_ab + score_a, 
    groups = "group_ab"
)
```

With 'full' model, we mean the total influence of `x`, including all of its 
interaction terms, on `y`.
For example, if our input formula was: 
`y ~ x * (group_ab + score_a)`. R rewrites this as follows: 
```{r, reformulate, echo = FALSE}
update.formula(y ~ x * (group_ab + score_a), . ~ .)
```
The variables that constitute the 'full' effect of x would be: 
`x + x:group_ab + x:score_a`. 

```{r plot_full, echo=FALSE, warning=FALSE, message=FALSE}
library(patchwork)
library(ggplot2)

full_df <- data.frame(
  citrate = tableX(web)[,'citrate'], 
  `cis-aconitate` = tableX(web)[,'cis-aconitate'],
  aconitase = tableY(web)[,'aconitase']
  )

ggplot(full_df) + 
  aes(x = aconitase, y = citrate) + 
  facet_wrap( ~ "citrate")  +

  ggplot(full_df) + 
  aes(x = aconitase, y = cis.aconitate) + 
  facet_wrap( ~ "cis-aconitate") +
  
  plot_layout(axes = 'collect', axis_titles = 'collect') &

  scale_y_continuous(limits = c(-2.5, 2.5)) &
  geom_point(shape = 21, fill = "#ffffbf") &
  theme_bw()
```

### compare to `lm()`. 

```{r lm-compare-full} 
```

# Differential association {#sec-da-2}

In order to assess differences in associations based on one or more variables 
(such as phenotype or treatment), we make use of the emergent and disjointed 
association paradigm introduced in the context of proportionality \citep{diff_prop,quinn2017propr,part_corr} and apply it outside of the simplex. 
Briefly, disjointed associations refer to the scenario where the \textit{slope} 
of an association is dependent on a variable. On the other hand, emergent 
associations refer to the scenario where the \textit{strength} of the scenario 
is dependent on a variable. 

## Disjointed associations
Disjointed associations describe the scenario where there is 
a detectable association in all cases, but the quality of that association 
differs. 

### Interpretation:
Disjointed associations describe the situation where there is a detectable association in all cases, but the quality of that association, the **slope**, changes in a manner explained by a term. We estimate this as interaction terms with `x`. In our example we have two such terms, namely `x:group_ab` and `x:score_a`. 

### Disjointed associations with categorical variables

```{r disj_cat, echo=FALSE} 
diff_df <- data.frame(
  group_ab      = metadata(web)$group_ab, 
  score_a_cat   = cut(metadata(web)$score_a, 
                      quantile(metadata(web)$score_a, probs = seq(0,1, 1/5)), 
                      include.lowest = TRUE
                      ),
  score_a       = metadata(web)$score_a,
  isocitrate    = tableX(web)[,'isocitrate'],
  ketoglutarate = tableX(web)[,'ketoglutarate'],
  isocitrate.dehydrogenase = tableY(web)[,'isocitrate dehydrogenase'],
  ketoglutarate.dehydrogenase = tableY(web)[,'ketoglutarate dehydrogenase'], 
  succinyl.CoA  = tableX(web)[,'succinyl-CoA'],
  succinate = tableX(web)[,'succinate'],
  succinyl.CoA.synthetase = tableY(web)[,'succinyl-CoA synthetase'],
  succinate.dehydrogenase = tableY(web)[,'succinate dehydrogenase']
  )

ggplot(diff_df) + 
  aes(x = isocitrate.dehydrogenase, y = isocitrate, fill = group_ab) + 
  facet_wrap(~group_ab) +

  geom_point(shape = 21, show.legend = FALSE) +
  scale_fill_manual(values = c("a" = "#d73027", 
                               "b" = "#2166ac")) + 
  theme_bw() + 
  coord_equal(ratio = 1)
```

### Disjointed associations with continuous variables

```{r disj_cont, echo=FALSE} 

  ggplot(diff_df) + 
  
  aes(x = ketoglutarate.dehydrogenase, y = ketoglutarate, fill = score_a) + 
  facet_grid(~score_a_cat) +
  
  geom_point(shape = 21, show.legend = FALSE) +
    scale_fill_gradient2(low = "#a50026", mid = "#ffffbf", high = "#313695", ) +

  theme_bw() + 
  coord_equal(ratio = 1)
```

### compare to `lm()` outcome.

```{r lm-disj} 

```

## Emergent association
Emergent associations describe the situation where the **strength** of an association is explained by a term interacting with x. We again estimate this as interaction terms with `x`. In our example we have two such terms, namely `x:group_ab` and `x:score_a`. 

### Emergent associations with categorical variables

```{r emerg_cat, echo=FALSE}

ggplot(diff_df) + 
  aes(x = succinyl.CoA.synthetase, y = succinyl.CoA, fill = group_ab) + 
  facet_wrap(~group_ab) +
 
  geom_point(shape = 21, show.legend = FALSE) +
   scale_fill_manual(values = c("a" = "#d73027", 
                               "b" = "#2166ac")) + 
  
  theme_bw() + 
  coord_equal(ratio = 1)
```

### Emergent associations with continuous variables

```{r emerg_cont, echo=FALSE}

ggplot(diff_df) + 
  aes(x = succinate.dehydrogenase, y = succinate, fill = score_a) +
  facet_grid(~score_a_cat) +

  geom_point(shape = 21, show.legend = FALSE) +
    scale_fill_gradient2(low = "#a50026", mid = "#ffffbf", high = "#313695") +

  theme_bw() + 
  coord_equal(ratio = 1)
```

# Repeated measures with `Error()` {#sec-da-3}

## Random slopes through `Error()`

```{r intro-error} 
# out <- anansi(AnansiWeb, formula = ~ treatment * temperature + Error(participant_id), metadata = metadata)
```

### compare to `aov( Error() )`.

```{r aov-error} 
```

### corresponding `lm()`

```{r lm-error} 
```

# Session info

```{r}
sessionInfo()
```
