---
title: "Association testing"
output:
    BiocStyle::html_document:
        toc: true
        toc_float: true
        toc_depth: 2
        number_sections: true
vignette: >
  %\VignetteIndexEntry{3. Differential associations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
body {
text-align: justify}
</style>

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(anansi)
```


# Overview {-}
A common challenge in integrative analysis of high-dimensional data sets is 
subsequent biological interpretation. Anansi addresses this challenge by only 
considering pairwise associations that are known to occur *a priori*. In order
to achieve this, we need to provide this relational information to anansi in the
form of a (bi)adjacency matrix. 

This vignette is divided by three sections:

* [Section 1](#sec-da-1) introduces the anansi framework and covers the *full 
model*.  
* [Section 2](#sec-da-2) Covers differential associations
* [Section 3](#sec-da-3) Discusses incorporating repeated measures through 
`Error()` notation. 

# Getting started though the full model {#sec-da-1}
Overview
full model estimates the effect of x on y.

## formula syntax. 

```{r formula-web-full} 
# setup anansiWeb
tableY <- matrix( 
rnorm(60*10), nrow = 60, dimnames = list(ids = 1:60, y = LETTERS[1:10] )
) 

tableX <- matrix( 
rnorm(60*20), nrow = 60, dimnames = list(ids = 1:60, x = letters[1:20] )
) 

anansiWeb <- weaveWeb( y ~ x, link = "none", 
                       tableY = tableY, tableX = tableX ) 
```

### link to adjacency matrix vignette

```{r link-terms} 
terms(anansiWeb) 
```

```{r formula-anansi-full} 
# out <- anansi(anansiWeb, formula = ~ treatment * temperature, 
# metadata = metadata)
```
Anansi supports arbitrarily complex linear models as well as longitudinal models
using the R formula syntax. 

full is the `x` part of `y ~ x * (a + b + c)`:

drop `x + x:a + x:b + x:c`. 

### compare to `lm()`. 

```{r lm-compare-full} 

```

# Differential association {#sec-da-2}

In order to assess differences in associations based on one or more variables 
(such as phenotype or treatment), we make use of the emergent and disjointed 
association paradigm introduced in the context of proportionality \citep{diff_prop,quinn2017propr,part_corr} and apply it outside of the simplex. 
Briefly, disjointed associations refer to the scenario where the \textit{slope} 
of an association is dependent on a variable. On the other hand, emergent 
associations refer to the scenario where the \textit{strength} of the scenario 
is dependent on a variable. See figure \@ref(fig:fig-diffab) for an illustrated 
example. 

An example of differential associations between hypothetical features Y and X. 
In both cases, phenotype C illustrates the differential association compared to 
phenotypes A \& B. Disjointed associations describe the scenario where there is 
a detectable association in all cases, but the quality of that association 
differs. Emergent associations describe the case where an association can be 
detected in one case but not in another. 

The features Y and X are from different data sets and differential associations 
can be expressed in the style of R classical linear models: 
$lm(Y \sim X \underline{\times Phenotype})$ and 
$lm(abs(residuals( lm(Y \sim X)) ) \sim \underline{Phenotype})$ for disjointed 
and emergent associations, respectively.

```{r} 
#| fig-diffab, fig.wide=TRUE, warning=FALSE,message=FALSE,echo=FALSE,
#| fig.cap = "An example of differential associations between hypothetical
#| features Y and X. In both cases, phenotype C illustrates the differential 
#| association compared to phenotypes A & B. Disjointed associations describe 
#| the scenario where there is a detectable association in all cases, but the 
#| quality of the association differs. Emergent associations describe the case 
#| where an association can be detected in one case but not in another." 

library(tidyverse)
library(patchwork)
library(tidyverse)

set.seed(1)
a = data.frame(Y  = c(1:10, (1:10)*3/2, 10:1) + runif(30, -1, 1),
               X  = rep(1:10,  3) + runif(30, -1, 1),
               Phenotype = rep(LETTERS[1:3], each = 10),
               plot = "Disjointed Association")

b = data.frame(Y  = c(1:10, (1:10)*3/2, (1:10)*5/4) + c(runif(20, -1, 1), runif(10, -7, 7)),
               X  = rep(1:10,  3)  + c(runif(20, -1, 1), runif(10, -7, 7)),
               Phenotype = rep(LETTERS[1:3], each = 10),
               plot = "Emergent Association")

  ggplot( rbind(a, b) ) +
  aes(x = X, y = Y, fill = Phenotype, colour = Phenotype) +
  geom_point(size = 3, shape = 21, colour = "black") +
  geom_smooth(method = "lm", se = F, linetype = "dashed", show.legend = F) +
  facet_wrap(~plot) +
  theme_bw() 

```

### ref `propr` paper.

## Disjointed associations
Overview

### Interpretation:
disjointed is term interaction with `x`, namely, drop `x:a`, `x:b`, `x:c`. 

### compare to `lm()` outcome.

```{r lm-disj} 

```

## Emergent association
    decide on action. 

# Repeated measures with `Error()` {#sec-da-3}

## Random slopes through `Error()`

```{r intro-error} 

# out <- anansi(anansiWeb, formula = ~ treatment * temperature + Error(participant_id), metadata = metadata)

```

### compare to `aov( Error() )`.

```{r aov-error} 

```

### corresponding `lm()`

```{r lm-error} 

```

# Session info

```{r}
sessionInfo()
```
